Toward \earlywarning\ detection of gravitational waves from compact binary coalescence

### 2011-08-12 conclusions.tex
Fact checking pass: updated flop/s table, corrected use of Nup and Ndown, stripped descriptions of pipeline as 'real-time'
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1339 ccdd05d5-56c1-461a-bf98-85f6876523b4

The \lloid{} method is as fast as standard \fft{} convolutions but allows for
nearly latency-free, real-time operation.  We anticipate requiring $\sim$100
modern multi-core computers to search for binary neutron stars using
coincident \GW\ data from a multi-detector network.  In the future, additional
computational savings could be acheived by conditionally reconstructing the
\SNR{} time-series only during times when a composite detection statistic
crosses a threshold~\citep{svd-compdetstat}.  However, the anticipated required
number of computers is well within the current computing capabilities of the
\LIGO{} Data Grid.

>>>>

The \lloid{} method is as fast as standard \fft{} convolutions but allows for
nearly latency-free, real-time operation.  We anticipate requiring $\sim$400
modern multi-core computers to search for binary neutron stars using
coincident \GW\ data from a multi-detector network.  In the future, additional
computational savings could be acheived by conditionally reconstructing the
\SNR{} time-series only during times when a composite detection statistic
crosses a threshold~\citep{svd-compdetstat}.  However, the anticipated required
number of computers is well within the current computing capabilities of the
\LIGO{} Data Grid.

### 2011-08-12 figures/diagram.tex
Converted diagram to eps
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1165 ccdd05d5-56c1-461a-bf98-85f6876523b4
lloid_paper: fix LLOID diagram generation
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1249 ccdd05d5-56c1-461a-bf98-85f6876523b4
Fact checking pass: updated flop/s table, corrected use of Nup and Ndown, stripped descriptions of pipeline as 'real-time'
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1339 ccdd05d5-56c1-461a-bf98-85f6876523b4

\documentclass{amsart}
\usepackage{tikz}
\usetikzlibrary{external,decorations.pathreplacing,matrix,fit,calc,chains,scopes}
\tikzexternalize[force remake=true]

@@

		column 2/.style={column sep=15mm},
		column 3/.style={column sep=3mm},
		column 4/.style={column sep=8mm},
		column 5/.style={column sep=15mm},
		column 6/.style={column sep=8mm},
		column 11/.style={column sep=8mm}

@@

		& \node[yshift=0.5cm](in){}; & \node[yshift=0.5cm](intee){}; & \node[yshift=0.5cm](u){}; & \node[yshift=0.5cm](utee){}; & \node[yshift=0.5cm](v){}; & \node[yshift=0.5cm](add0){}; & \node[yshift=0.5cm](add1){}; & \node[yshift=0.5cm](add2){}; & & \node[yshift=0.5cm](add3){}; & \node[yshift=0.5cm](out){}; \\
		& & & \node[draw] (u00) {}; & \node[junction] (u00tee) {}; & \node[adder] (v00) {}; & \node[adder] (add00) {}; & & & & & \node[signal, label=right:{$\rho_0^0$}] (out00) {}; \\
		& & & \node[draw] (u10) {}; & \node[junction] (u10tee) {}; & \node[adder] (v10) {}; & & \node[adder] (add10) {}; & & & & \node[signal, label=right:{$\rho_1^0$}] (out10) {}; \\
		\node[signal, label=left:{$x$}] (x) {}; & \node[junction] (in0) {}; & \node[junction] (in0tee) {}; & & & \node[adder] (v20) {}; & & & \node[adder] (add20) {}; & & & \node[signal, label=right:{$\rho_2^0$}] (out20) {}; \\
		& & & & & \node[vdots]{$\colon$}; & & & & & & \node[vdots]{$\colon$}; \\
		& & & \node[draw] (u20) {}; & \node[junction] (u20tee) {}; & \node[adder] (v30) {}; & & & & & \node[adder] (add30) {}; & \node[signal, label=right:{$\rho_{M-1}^0$}] (out30) {}; \\
		& & & & & & & & \node[minimum height=6mm](delay0){}; & & & \\
		& & & & & & \node[resample] (resample01) {\tiny$\uparrow$}; & \node[resample] (resample11) {\tiny$\uparrow$}; & \node[resample] (resample21) {\tiny$\uparrow$}; & \node{$\cdot \cdot$}; & \node[resample] (resample31) {\tiny$\uparrow$}; \\

@@

		& & & \node[draw] (u01) {}; & \node[junction] (u01tee) {}; & \node[adder] (v01) {}; & \node[adder] (add01) {}; & & & & & \node[signal, label=right:{$\rho_0^1$}] (out01) {}; \\
		& & & \node[draw] (u11) {}; & \node[junction] (u11tee) {}; & \node[adder] (v11) {}; & & \node[adder] (add11) {}; & & & & \node[signal, label=right:{$\rho_1^1$}] (out11) {}; \\
		& \node[resample] (in1) {\tiny$\downarrow$}; & \node[junction] (in1tee) {}; & & & \node[adder] (v21) {}; & & & \node[adder] (add21) {}; & & & \node[signal, label=right:{$\rho_2^1$}] (out21) {}; \\
		& & & & & \node[vdots]{$\colon$}; & & & & & & \node[vdots]{$\colon$}; \\
		& & & \node[draw] (u21) {}; & \node[junction] (u21tee) {}; & \node[adder] (v31) {}; & & & & & \node[adder] (add31) {}; & \node[signal, label=right:{$\rho_{M-1}^1$}] (out31) {}; \\
		& & & & & & & & \node[minimum height=6mm](delay1){}; & & & \\
		& & & & & & \node[resample] (resample02) {\tiny$\uparrow$}; & \node[resample] (resample12) {\tiny$\uparrow$}; & \node[resample] (resample22) {\tiny$\uparrow$}; & \node{$\cdot \cdot$}; & \node[resample] (resample32) {\tiny$\uparrow$}; \\

@@

		& \node[minimum size=0pt, text height=8pt, anchor=base] (invdots) {$\vdots$}; & & \node[minimum size=0pt, text height=0pt, anchor=base] {$\vdots$}; & \node[minimum size=0pt, text height=8pt, anchor=base] {$\vdots$}; & \node[minimum size=0pt, text height=8pt, anchor=base] {$\vdots$}; & \node[minimum height=8pt]{}; & & & & & \node[minimum size=0pt, text height=0pt, anchor=base] {$\vdots$}; \\
		& & & & & & \node[minimum size=0pt, text height=8pt] (vdots0) {$\vdots$}; & \node[minimum size=0pt, text height=8pt] (vdots1) {$\vdots$}; & \node[minimum size=0pt, text height=8pt] (vdots2) {$\vdots$}; & & \node[minimum size=0pt, text height=8pt] (vdots3) {$\vdots$}; \\

@@

		& & & \node[draw] (u02) {}; & \node[junction] (u02tee) {}; & \node[adder] (v02) {}; & \node[junction] (add02) {}; & & & & & \node[signal, label=right:{$\rho_0^{S-1}$}] (out02) {}; \\
		& & & \node[draw] (u12) {}; & \node[junction] (u12tee) {}; & \node[adder] (v12) {}; & & \node[junction] (add12) {}; & & & & \node[signal, label=right:{$\rho_1^{S-1}$}] (out12) {}; \\
		& \node[junction] (in2) {}; & \node[junction] (in2tee) {}; & & & \node[adder] (v22) {}; & & & \node[junction] (add22) {}; & & & \node[signal, label=right:{$\rho_2^{S-1}$}] (out22) {}; \\
		& & & & & \node[vdots]{$\colon$}; & & & & & & \node[vdots]{$\colon$}; \\
		& & & \node[draw] (u22) {}; & \node[junction] (u22tee) {}; & \node[adder] (v32) {}; & & & & & \node[junction] (add32) {}; & \node[signal, label=right:{$\rho_{M-1}^{S-1}$}] (out32) {}; \\

@@

	\draw (in0) -- (in0tee) node[midway,above] {\small 4096 Hz};
	\draw (in1) -- (in1tee) node[midway,above] {\small 512 Hz};
	\draw (in2) -- (in2tee) node[midway,above] {\small 32 Hz};

@@

	\node[draw,fill=white,minimum width=22mm,xshift=1mm,label=right:{\small delay}] at (delay0) {\tiny$z^{-t^1 f^0}$};
	\node[draw,fill=white,minimum width=22mm,xshift=1mm,label=right:{\small delay}] at (delay1) {\tiny$z^{-(t^2-t^1) f^1}$};
	\draw[->] (in0) -- (in1);
	\draw (in1) -- (invdots);
	\draw[->] (invdots) -- (in2);
	\draw[->] (x) -- (in0);

	\draw[decorate, decoration={brace, amplitude=4pt}] ($(in)+(-4mm,0)$) -- ($(in)+(4mm,0)$) node[midway,above=2mm,align=center] {Decimation \\ of input};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(intee)+(-2mm,0)$) -- ($(u)+(2mm,0)$) node[midway,above=2mm,align=center] {Orthogonal \\ \textsc{fir} filters};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(utee)+(-2mm,0)$) -- ($(v)+(2mm,0)$) node[midway,above=2mm,align=center] {Reconstruction \\ matrices};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(add0)+(-2mm,0)$) -- ($(add3)+(2mm,0)$) node[midway,above=2mm,align=center] {Interpolation and \\ \textsc{snr} accumulation};
\end{tikzpicture}

>>>>

\documentclass{amsart}
\usepackage{tikz}
\usetikzlibrary{external,decorations.pathreplacing,matrix,fit,calc,chains,scopes}
\tikzexternalize[force remake=true]
\tikzset{external/system call={pdflatex \tikzexternalcheckshellescape -halt-on-error -interaction=batchmode -jobname "\image" "\texsource"; pdftops -eps "\image".pdf "\image".eps}}

@@

		column 2/.style={column sep=2mm},
		column 3/.style={column sep=12mm},
		column 4/.style={column sep=3mm},
		column 5/.style={column sep=8mm},
		column 6/.style={column sep=15mm},
		column 7/.style={column sep=8mm},
		column 12/.style={column sep=8mm}

@@

		& \node[yshift=0.5cm](in){}; & & \node[yshift=0.5cm](intee){}; & \node[yshift=0.5cm](u){}; & \node[yshift=0.5cm](utee){}; & \node[yshift=0.5cm](v){}; & \node[yshift=0.5cm](add0){}; & \node[yshift=0.5cm](add1){}; & \node[yshift=0.5cm](add2){}; & & \node[yshift=0.5cm](add3){}; & \node[yshift=0.5cm](out){}; \\
		& & & & \node[draw] (u00) {}; & \node[junction] (u00tee) {}; & \node[adder] (v00) {}; & \node[adder] (add00) {}; & & & & & \node[signal, label=right:{$\rho_0^0$}] (out00) {}; \\
		& & & & \node[draw] (u10) {}; & \node[junction] (u10tee) {}; & \node[adder] (v10) {}; & & \node[adder] (add10) {}; & & & & \node[signal, label=right:{$\rho_1^0$}] (out10) {}; \\
		\node[signal, label=left:{$x$}] (x) {}; & \node[junction] (in0) {}; & & \node[junction] (in0tee) {}; & & & \node[adder] (v20) {}; & & & \node[adder] (add20) {}; & & & \node[signal, label=right:{$\rho_2^0$}] (out20) {}; \\
		& & & & & & \node[vdots]{$\colon$}; & & & & & & \node[vdots]{$\colon$}; \\
		& & & & \node[draw] (u20) {}; & \node[junction] (u20tee) {}; & \node[adder] (v30) {}; & & & & & \node[adder] (add30) {}; & \node[signal, label=right:{$\rho_{M-1}^0$}] (out30) {}; \\
		& & & & & & & & & \node[minimum height=6mm](delay0){}; & & & \\
		& & & & & & & \node[resample] (resample01) {\tiny$\uparrow$}; & \node[resample] (resample11) {\tiny$\uparrow$}; & \node[resample] (resample21) {\tiny$\uparrow$}; & \node{$\cdot \cdot$}; & \node[resample] (resample31) {\tiny$\uparrow$}; \\

@@

		& & & & \node[draw] (u01) {}; & \node[junction] (u01tee) {}; & \node[adder] (v01) {}; & \node[adder] (add01) {}; & & & & & \node[signal, label=right:{$\rho_0^1$}] (out01) {}; \\
		& & & & \node[draw] (u11) {}; & \node[junction] (u11tee) {}; & \node[adder] (v11) {}; & & \node[adder] (add11) {}; & & & & \node[signal, label=right:{$\rho_1^1$}] (out11) {}; \\
		& \node[junction] (in1) {}; & \node[resample] (in1downsample) {\tiny$\downarrow$}; & \node[junction] (in1tee) {}; & & & \node[adder] (v21) {}; & & & \node[adder] (add21) {}; & & & \node[signal, label=right:{$\rho_2^1$}] (out21) {}; \\
		& & & & & & \node[vdots]{$\colon$}; & & & & & & \node[vdots]{$\colon$}; \\
		& & & & \node[draw] (u21) {}; & \node[junction] (u21tee) {}; & \node[adder] (v31) {}; & & & & & \node[adder] (add31) {}; & \node[signal, label=right:{$\rho_{M-1}^1$}] (out31) {}; \\
		& & & & & & & & & \node[minimum height=6mm](delay1){}; & & & \\
		& & & & & & & \node[resample] (resample02) {\tiny$\uparrow$}; & \node[resample] (resample12) {\tiny$\uparrow$}; & \node[resample] (resample22) {\tiny$\uparrow$}; & \node{$\cdot \cdot$}; & \node[resample] (resample32) {\tiny$\uparrow$}; \\

@@

		& \node[minimum size=0pt, text height=8pt, anchor=base] (invdots) {$\vdots$}; & & & \node[minimum size=0pt, text height=0pt, anchor=base] {$\vdots$}; & \node[minimum size=0pt, text height=8pt, anchor=base] {$\vdots$}; & \node[minimum size=0pt, text height=8pt, anchor=base] {$\vdots$}; & \node[minimum height=8pt]{}; & & & & & \node[minimum size=0pt, text height=0pt, anchor=base] {$\vdots$}; \\
		& & & & & & & \node[minimum size=0pt, text height=8pt] (vdots0) {$\vdots$}; & \node[minimum size=0pt, text height=8pt] (vdots1) {$\vdots$}; & \node[minimum size=0pt, text height=8pt] (vdots2) {$\vdots$}; & & \node[minimum size=0pt, text height=8pt] (vdots3) {$\vdots$}; \\

@@

		& & & & \node[draw] (u02) {}; & \node[junction] (u02tee) {}; & \node[adder] (v02) {}; & \node[junction] (add02) {}; & & & & & \node[signal, label=right:{$\rho_0^{S-1}$}] (out02) {}; \\
		& & & & \node[draw] (u12) {}; & \node[junction] (u12tee) {}; & \node[adder] (v12) {}; & & \node[junction] (add12) {}; & & & & \node[signal, label=right:{$\rho_1^{S-1}$}] (out12) {}; \\
		& \node[] (in2) {}; & \node[resample] (in2downsample) {\tiny$\downarrow$}; & \node[junction] (in2tee) {}; & & & \node[adder] (v22) {}; & & & \node[junction] (add22) {}; & & & \node[signal, label=right:{$\rho_2^{S-1}$}] (out22) {}; \\
		& & & & & & \node[vdots]{$\colon$}; & & & & & & \node[vdots]{$\colon$}; \\
		& & & & \node[draw] (u22) {}; & \node[junction] (u22tee) {}; & \node[adder] (v32) {}; & & & & & \node[junction] (add32) {}; & \node[signal, label=right:{$\rho_{M-1}^{S-1}$}] (out32) {}; \\

@@

	\draw (in0) -- (in0tee) node[midway,above] {\small 4096 Hz};
	\draw[->] (in1) -- (in1downsample);
	\draw (in1downsample) -- (in1tee) node[pos=0.4,above] {\small 512 Hz};
	\draw (in2downsample) -- (in2tee) node[pos=0.4,above] {\small 32 Hz};

@@

	\node[draw,fill=white,minimum width=22mm,xshift=1mm,label=right:{\small delay}] at (delay0) {\tiny$z^{-t^1 f^0}$};
	\node[draw,fill=white,minimum width=22mm,xshift=1mm,label=right:{\small delay}] at (delay1) {\tiny$z^{-(t^2-t^1) f^1}$};
	\draw[->] (in0) -- (in1);
	\draw (in1) -- (invdots);
	\draw[->] (invdots) |- (in2downsample);
	\draw[->] (x) -- (in0);

	\draw[decorate, decoration={brace, amplitude=4pt}] ($(in)+(-4mm,0)$) -- ($(in)+(10mm,0)$) node[midway,above=2mm,align=center] {Decimation \\ of input};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(intee)+(-2mm,0)$) -- ($(u)+(2mm,0)$) node[midway,above=2mm,align=center] {Orthogonal \\ \textsc{fir} filters};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(utee)+(-2mm,0)$) -- ($(v)+(2mm,0)$) node[midway,above=2mm,align=center] {Reconstruction \\ matrices};
	\draw[decorate, decoration={brace, amplitude=4pt}] ($(add0)+(-2mm,0)$) -- ($(add3)+(2mm,0)$) node[midway,above=2mm,align=center] {Interpolation and \\ \textsc{snr} accumulation};
\end{tikzpicture}

### 2011-08-12 implementation.tex
Fact checking pass: updated flop/s table, corrected use of Nup and Ndown, stripped descriptions of pipeline as 'real-time'
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1339 ccdd05d5-56c1-461a-bf98-85f6876523b4

The \lloid\ algorithm is amenable to nearly latency free, real-time implementation.
However, such a near real-time search pipeline would likely require integration directly into
the data acquisition and storage system of the \LIGO\ observatories.  A slightly more 
modest goal is to leverage existing low-latency, but not real-time, signal processing
software in order to implement the \lloid\ algorithm.

@@

The whitened detector data is reduced to successively
lower sample rates by decimation. Decimation involves applying an anti-aliasing
filter to the data, and then down-sampling by deleting samples.  We use a
192-tap \fir\ decimator provided by \gstreamer{}'s {\tt audioresample}
element.  The detector data is provided at every power-of-two sample rate
required by the time slices described in \eqref{eq:time-slices}.  At this point,
the \gstreamer\ pipeline splits into several parallel branches to handle each
time slice.

>>>>

The \lloid\ algorithm is amenable to extremely low latency implementation.
The lowest possibly latency could be achieved by integration directly into
the data acquisition and storage system of the \LIGO\ observatories.  A slightly more 
modest goal is to leverage existing signal processing software in order to implement
the \lloid\ algorithm.

@@

The whitened detector data is reduced to successively
lower sample rates by decimation. Decimation involves applying an anti-aliasing
filter to the data, and then down-sampling by deleting samples.  We use GStreamer's
stock {\tt audioresample} element, which provides a FIR decimation filter with
a Kaiser-windowed sinc function kernel.

The detector data is provided at every power-of-two sample rate
required by the time slices described in \eqref{eq:time-slices}.  At this point,
the \gstreamer\ pipeline splits into several parallel branches to handle each
time slice.

### 2011-08-12 method.tex
Fact checking pass: updated flop/s table, corrected use of Nup and Ndown, stripped descriptions of pipeline as 'real-time'
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@1339 ccdd05d5-56c1-461a-bf98-85f6876523b4

Equation~(\ref{eq:SNRTD}) can be implemented in the time-domain as bank of
finite impulse response (\fir) filters, requiring $\mathcal O(\numtmps
\tmpsamps)$ floating point operations per sample.  However, it is typically
much more computationally efficient to use the convolution theorem and the
Fast Fourier Transform (\fft) to implement fast convolution in the frequency
domain, requiring only
$\mathcal O(\numtmps \lg \tmpsamps)$ operations per sample but incurring
a latency of $\mathcal O(\tmpsamps)$ samples.

@@

Here we describe a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We give a nearly zero-latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which by contrast requires a significant latency due to the
inherent acausality of the Fourier transform.  Our method, called \lloid{}
(Low Latency Online Inspiral Detection),
involves two transformations of the templates that produce a
network of orthogonal filters that is far more computationally
efficient than the original bank of matched filters.

@@

As noted previously, the template banks used in inspiral searches are by design
highly correlated.  \citet{Cannon:2010p10398} showed that applying the \SVD\
to inspiral template banks greatly reduces the number of filters required to achieve a
particular minimal match.  A similar technique can be applied to the time-sliced
templates as defined in equation~\ref{eq:time-sliced-templates} above.  The \SVD\
is a matrix factorization that takes the form
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
where $u_l^s[k]$ are orthonormal \emph{basis templates} related to the original
time-sliced templates through the \emph{reconstruction matrix}, $v_{il}^s\sigma_l^s$.
The expectation value of the fractional loss in \SNR\ is the \SVD\ tolerance, given by
%
\begin{equation*}
\left[ \sum_{l=0}^{L^s-1} \left( \sigma_l^s \right)^2 \right]\left[ \sum_{l=0}^{M-1} \left( \sigma_l^s \right)^2 \right]^{-1},
\end{equation*}
%
determined by the number $\numsvdtmps$ of basis templates that are kept in
the approximation.  The authors of \citet{Cannon:2010p10398}
showed that highly accurate approximations of inspiral template banks could be
achieved with few basis templates.  We find that when combined with the
time slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on the
rank reduction demonstrated in \citep{Cannon:2010p10398} by nearly an order
of magnitude.  In the next section we describe how we form our early-warning
detection statistic using the time slice decomposition and the \SVD.

@@

First, the sample rate of the detector data must be decimated to match sample rates
with each of the time slices.  We will denote the decimated detector data streams using
a superscript $^s$ to indicate the time slices to which they correspond, and the
operator $H^\shortdownarrow$ will represent the appropriate decimation filter:
\begin{equation*}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k].
\end{equation*}
We shall use the symbol $H^\shortuparrow$ to represent an interpolation filter,
\begin{equation*}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k].
\end{equation*}
Ultimately, the latency of the entire \lloid\ algorithm is set by the decimation
and interpolation filters because they are generally time-symmetric.  As a result,
\lloid\ is very slightly acausal with an intrinsic latency of
$\left(\max \left( N^\shortdownarrow, N^\shortuparrow \right) / 2 f^0 \right)$~seconds.

@@

We now examine the computational cost scaling of the conventional \TD{} or
\FD{} matched filter procedure as compared with \lloid{}.  For convenience,
Table~\ref{tab:recap} provides a review of the notation that we will need in
this section.
%
% symbols used in FLOPs calculations table
%
\begin{table}
\caption{\label{tab:recap}Notation used to describe filters.}
\begin{center}
\begin{tabular}{ll}
\tableline\tableline
& Definition \\
\tableline
$f^s$		& sample rate in time slice $s$ \\
\numtmps		& num. templates \\
\tmpsamps	& num. samples per template \\
\numslices	& num. time slices \\
\numsvdtmps	& num. basis templates in time slice $s$ \\
\slicessamps	& num. samples in decimated time slice $s$\\
$N^\shortdownarrow$ & num. coefficients in decimation filter \\
$N^\shortuparrow$ & num. coefficients in interpolation filter \\
\tableline
\end{tabular}
\end{center}
\end{table}

@@

\subsubsection{\lloid\ method}

@@

The computational cost of the decimation of the detector data is a little bit more subtle.  Decimation is achieved by applying an \fir\ anti-aliasing filter and then downsampling, or deleting samples in order to reduce the sample rate from $f^{s-1}$ to $f^s$.  Naively, an anti-aliasing filter with $N^\shortdownarrow$ coefficients should demand $2 N^\shortdownarrow f^{s-1}$ \flops.  However, it is necessary to evaluate the anti-aliasing filter only for the fraction $f^s / f^{s-1}$ of the samples that will not be deleted.  Consequently, an efficient decimator that requires only $2 N^\shortdownarrow f^{s-1} \cdot \left( f^s / f^{s-1} \right) = 2 N^\shortdownarrow f^s$ \flops\ is possible.

The story is similar for the interpolation filters used to match the sample rates of the partial \SNR{} streams.  Interpolation of a data stream from a sample rate $f^s$ to $f^{s-1}$ consists of inserting zeros between the samples of the original stream, and then applying a low-pass filter with $N^\shortdownarrow$ coefficients.  The low-pass filter requires $2 M N^\shortdownarrow f^{s-1}$ \flops.  However, by taking advantage of the fact that by construction a fraction $f^{s-1}/f$ of the samples are zero, it is possible to build an efficient interpolator that requires only $M N^\shortuparrow f^{s-1} \cdot \left( f^s / f^{s-1} \right) = 2 M N^\shortuparrow f^s$ \flops.

Taking into account the decimation of the detector data, the orthogonal \fir\ filters, the reconstruction of the time-sliced templates, the interpolation of \SNR\ from previous time slices, and the accumulation of \SNR, in total the \lloid\ algorithm requires
\begin{multline*}
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s \\ + 2\sum_{\mathclap{f \in \{f^s\}}} \left( N^\shortdownarrow + \numtmps N^\shortuparrow \right) f^s
\end{multline*}
\flops.  The second sum is carried out over the set of distinct sample rates rather than over the time slices themselves.

@@

>>>>

Equation~(\ref{eq:SNRTD}) can be implemented in the time-domain as a bank of
finite impulse response (\fir) filters, requiring $\mathcal O(\numtmps
\tmpsamps)$ floating point operations per sample.  However, it is typically
much more computationally efficient to use the convolution theorem and the
Fast Fourier Transform (\fft) to implement fast convolution in the frequency
domain, requiring only
$\mathcal O(\numtmps \lg \tmpsamps)$ operations per sample but incurring
a latency of $\mathcal O(\tmpsamps)$ samples.

@@

Here we describe a method for reducing the computational cost of a \TD\ search
for compact binary coalescence.  We give an extremely low latency algorithm
that competes in terms of floating point operations per second with the
conventional \FD\ method, which by contrast requires a significant latency due to the
inherent acausality of the Fourier transform.  Our method, called \lloid{}
(Low Latency Online Inspiral Detection),
involves two transformations of the templates that produce a
network of orthogonal filters that is far more computationally
efficient than the original bank of matched filters.

@@

As noted previously, the template banks used in inspiral searches are by design
highly correlated.  \citet{Cannon:2010p10398} showed that applying the \SVD\
to inspiral template banks greatly reduces the number of filters required to achieve a
particular minimal match.  A similar technique can be applied to the time-sliced
templates as defined in equation~\ref{eq:time-sliced-templates} above.  The \SVD\
is a matrix factorization that takes the form
%
\begin{equation}
h_i^s[k] = \sum_{\mathclap{l=0}}^{\mathclap{M-1}} v_{il}^s \sigma_l^s u_l^s[k] \approx \sum_{\mathclap{l=0}}^{\mathclap{L^s-1}} v_{il}^s \sigma_l^s u_l^s[k].
\label{eq:svddecomp}
\end{equation}
where $u_l^s[k]$ are orthonormal \emph{basis templates} related to the original
time-sliced templates through the \emph{reconstruction matrix}, $v_{il}^s\sigma_l^s$.
The expectation value of the fractional loss in \SNR\ is the \SVD\ tolerance, given by
%
\begin{equation*}
\left[ \sum_{l=0}^{L^s-1} \left( \sigma_l^s \right)^2 \right]\left[ \sum_{l=0}^{M-1} \left( \sigma_l^s \right)^2 \right]^{-1},
\end{equation*}
%
determined by the number $\numsvdtmps$ of basis templates that are kept in
the approximation.  The authors of \citet{Cannon:2010p10398}
showed that highly accurate approximations of inspiral template banks could be
achieved with few basis templates.  We find that when combined with the
time slice decomposition, the number of basis templates \numsvdtmps\ is much
smaller than the original number of templates \numtmps\ and improves on the
rank reduction demonstrated in \citep{Cannon:2010p10398} by nearly an order
of magnitude.

Because the sets of filters from each time slice form orthogonal subspaces, and the basis
basis filters within a given time slice are mutually orthogonal, the set of all basis filters
from all time slices forms an orthogonal basis spanning the original templates.

In the next section we describe how we form our early-warning
detection statistic using the time slice decomposition and the \SVD.

@@

First, the sample rate of the detector data must be decimated to match sample rates
with each of the time slices.  We will denote the decimated detector data streams using
a superscript $^s$ to indicate the time slices to which they correspond, and the
operator $H^\shortdownarrow$ will represent the appropriate decimation filter:
\begin{equation*}
\label{eq:decomp}
	x^{s+1}[k] = \left( H^\shortdownarrow x^s\right)[k].
\end{equation*}
We shall use the symbol $H^\shortuparrow$ to represent an interpolation filter,
\begin{equation*}
	x^{s}[k] = \left( H^\shortuparrow x^{s+1}\right)[k].
\end{equation*}
Ultimately, the latency of the entire \lloid\ algorithm is set by the decimation
and interpolation filters because they are generally time-symmetric and slightly
acausal.

@@

We now examine the computational cost scaling of the conventional \TD{} or
\FD{} matched filter procedure as compared with \lloid{}.  For convenience,
Table~\ref{tab:recap} provides a review of the notation that we will need in
this section.
%
% symbols used in FLOPs calculations table
%
\begin{table}
\caption{\label{tab:recap}Notation used to describe filters.}
\begin{center}
\begin{tabular}{ll}
\tableline\tableline
& Definition \\
\tableline
$f^s$		& sample rate in time slice $s$ \\
\numtmps		& num. templates \\
\tmpsamps	& num. samples per template \\
\numslices	& num. time slices \\
\numsvdtmps	& num. basis templates in time slice $s$ \\
\slicessamps	& num. samples in decimated time slice $s$\\
$N^\shortdownarrow$ & length of decimation filter \\
$N^\shortuparrow$ & length of interpolation filter \\
\tableline
\end{tabular}
\end{center}
\end{table}

@@

By adjusting the \fft\ block size, it is possible to achieve low latency with FD
convolution, but the computational cost grows rapidly as the latency in samples
$(D-N$) decreases.  It is easy to show that in the limit of many templates and
long templates, $M, \lg N \gg 1$, the computational cost scales as
%
$$
\left(1 + \frac{\textrm{template length}}{\textrm{latency}}\right)
$$
per template per $2\lg N$.

\subsubsection{\label{sec:lloid-method}\lloid\ method}

@@

The computational cost of the decimation of the detector data is a little bit more subtle.  Decimation is achieved by applying an \fir\ anti-aliasing filter and then downsampling, or deleting samples in order to reduce the sample rate from $f^{s-1}$ to $f^s$.  Naively, an anti-aliasing filter with $(f^{s-1} / f^s) N^\shortdownarrow$ coefficients should demand $2 N^\shortdownarrow (f^{s-1})^2 / f^s$ \flops.  However, it is necessary to evaluate the anti-aliasing filter only for the fraction $f^s / f^{s-1}$ of the samples that will not be deleted.  Consequently, an efficient decimator that requires only $2 N^\shortdownarrow f^{s-1}$ \flops\ is possible.

The story is similar for the interpolation filters used to match the sample rates of the partial \SNR{} streams.  Interpolation of a data stream from a sample rate $f^s$ to $f^{s-1}$ consists of inserting zeros between the samples of the original stream, and then applying a low-pass filter with $(f^{s-1} / f^s) N^\shortdownarrow$ coefficients.  The low-pass filter requires $2 M N^\shortdownarrow (f^{s-1})^2 / f^s$ \flops.  However, by taking advantage of the fact that by construction a fraction $f^s / f^{s-1}$ of the samples are zero, it is possible to build an efficient interpolator that requires only $M N^\shortuparrow f^{s-1}$ \flops.

Taking into account the decimation of the detector data, the orthogonal \fir\ filters, the reconstruction of the time-sliced templates, the interpolation of \SNR\ from previous time slices, and the accumulation of \SNR, in total the \lloid\ algorithm requires
\begin{multline*}
\sum_{\mathclap{s=0}}^{\mathclap{S-1}} \left( 2 \slicessamps \numsvdtmps + 2 \numtmps \numsvdtmps + \numtmps \right) f^s \\ + 2\sum_{\mathclap{f^s \in \{f^k \, : \, 0 < k < S\}}} \left( N^\shortdownarrow f^0 + \numtmps N^\shortuparrow f^{s-1}\right)
\end{multline*}
\flops.  The second sum is carried out over the set of distinct sample rates (except for the base sample rate) rather than over the time slices themselves, as we have found that it is sometimes desirable to place multiple adjacent time slices at the same sample rate in order to keep the size of matrices that enter the singular value decomposition manageable.  Here we have assumed that the decimation filters are connected in parallel, converting from the base sample rate $f^0$ to each of the time slice sample rates $f^1$, $f^2$, $\dots$, and that the interpolation filters are connected in cascade fashion with each interpolation filter stepping from the sample rate of one time slice to the next.

@@

The latency of the \lloid\ method is entirely due to the acausality of the interpolation and decimation filters.  The latency is exactly
$$
\frac{1}{2} \left( N^\shortdownarrow / f^{S-1} + \sum_{\mathclap{f^s \in \{f^k \, : \, 0 < k < S\}}} N^\shortuparrow / f^s \right) \textrm{ seconds.}
$$
The first term is due to the decimation filter that converts from the base sample rate $f^0$ to the lowest sample rate $f^{S-1}$ and the sum that comprises the second term is due to the latency of all of the interpolation filters connected in series.
